---
title: "Tennis davis Cup"
author: "Martí Casals & Jordi Cortes"
date: "`r Sys.Date()`"
output:
  prettydoc::html_pretty:
    theme: cayman
    highlight: github
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, fig.width = 10, fig.height = 7)
data_file <- '../Data/data_2023-03-30.Rdata'
```

## Introduction

### Data

The clean database is in the file: 

- __*`r data_file`*__.


```{r warning=FALSE, message=FALSE}
library(tidyverse)
library(forcats)
library(compareGroups)
library(psych)
library(car)
library(summarytools)
library(party)
library(emmeans)
library(multcomp)
library(effects)
library(AER)
library(knitr)
library(epiR)
library(SmartEDA)
library(data.table)
library(epitools)
library(patchwork) 

load(data_file)
dades <- dades_R; rm(dades_R)
n_WO  <- sum(dades$sum_score==0) # Walk Overs
```

### Additional debugging

- **Walk overs** (n = `r n_WO`) are removed.
- The average age variable of the players is constructed
- Fixed some factors


```{r}
##-- Mean age
dades$mean_age     <- dades$sum_age/2
dades$mean_age_cat <- round(dades$mean_age)

##-- Retirement as numeric
dades$Retirement_num <- as.numeric(dades$Retirement) - 1

##-- Arrange categorical variables
dades$best_of      <- factor(ifelse(dades$best_of=='5 sets','5sets','3sets'))
dades$round_cat    <- factor(ifelse(dades$round %in% c("QF","SF","F"),
                                    "Final","Preliminar"),
                             levels=c('Preliminar','Final'))
dades$mean_age_ind <- factor(ifelse(dades$mean_age>=27,">27","<27"))
dades$year_cat     <- factor(ifelse(dades$year<2010,"<2010",">=2010"),
                             levels=c('<2010','>=2010'))
dades$year_center  <- dades$year - median(dades$year)

##-- Remove walk overs (n=11)
dades$WO <- dades$sum_score==0
n_WO     <- sum(dades$WO)
dades    <- dades %>% filter(!WO)
```

First registers of data.

```{r}
head(dades,3)
```


# Descriptive analysis

## Covariates

```{r results='asis', message=FALSE, warning=FALSE}
VAR_ADJUST <- c('surface','best_of','group','year','sum_age','dif_age','match_num')
dfSummary(dades[,VAR_ADJUST], 
          style        = "grid",
          plain.ascii  = FALSE,
          graph.magnif = 0.75, 
          tmp.img.dir  = "./tmp")
```

## Retirements

```{r}
p_ret <- prop.table(table(dades$Retirement))[2]
```

The proportion of withdrawals in our sample is `r round(p_ret,3)`. That is, `r 100*round(p_ret,3)`%. 

```{r}
knitr::kable(as.data.frame(table(dades$Retirement)),
      col.names= c("Retirements", "Number"),
      align= "cc")
```

Retirements by group:

```{r}
with(dades,table(group,Retirement))
```

## Bivariate descriptive analysis

```{r warning=FALSE}
dades$surface <- relevel(dades$surface,ref="Carpet")
dades$group   <- relevel(dades$group,  ref="No Mundial 1")
dades$round   <- factor(dades$round, 
                        levels=c('PO','GS','SR','R1','R2','R3','QF','SF','F'))


res2  <- compareGroups(Retirement~surface+best_of+group+round_cat+match_num+sum_score+mean_age,
                       data=dades,byrow = TRUE)
summary(res2)
restab1 <- createTable(res2,show.n=TRUE,show.ratio=TRUE,type=2)
restab1 <- createTable(res2,show.n=FALSE,show.ratio=FALSE,show.p.overall = FALSE,type=2)
restab1
export2word(restab1,file = "Table1.docx")
```

```{r warning=FALSE}
#### Incidence per 1000 games
dd <- as.data.table(dades)[,.(retirements=sum(Retirement=='Yes'),games=sum(sum_score)),by=year]
pe <-pois.exact(x = dd$retirements,pt = dd$games)

per_param <- 1000
dd$incidence     <- per_param*pe[,'rate']
dd$incidence_low <- per_param*pe[,'lower']
dd$incidence_upp <- per_param*pe[,'upper']

ggplot(dd,aes(x=year,y=incidence)) + 
  geom_line (linewidth=1, col='darkblue') +
  geom_point(col='darkblue', size=3) +
  ylim(0,max(dd$incidence)) +
  xlim(2000,2020) +
  labs(x="Year", 
       y="Retirement Incidence Rate per 1,000 games",
       title="Davis Cup Incidence of Retirements by year") +
  theme(axis.title = element_text(face = 'bold'),
        axis.text  = element_text(face = 'bold'))

inc_rate_game_overall <- 1000*sum(dd$retirements)/sum(dd$games)
inc_rate_game_overall_per_1 <- pois.exact(x = sum(dd$retirements),pt = sum(dd$games))
inc_rate_game_overall_per_1000 <- inc_rate_game_overall_per_1
inc_rate_game_overall_per_1000[3:5] <- 1000*inc_rate_game_overall_per_1000[3:5]
inc_rate_game_overall_per_1000

#### Epidemiological measures

# Option 1:  Base R - tapply for SUM of people who RETIRED by SURFACE and the SUM of FOLLOW-UP TIME by SURFACE 
events <- tapply(dades$Retirement=="Yes", dades$surface, sum) 
persontime <- tapply(dades$sum_score, dades$surface, sum)
# make a 2 by 2 table, remembering to make the unexposed group the 2nd row 
tab_surface <- cbind(events, persontime) 
tab_surface <- tab_surface[c(2,1),]
epi.2by2(tab_surface, method = "cohort.time")

# Option 2: Via Tydiverse
retirement_rate <- dades %>% 
  group_by(surface) %>% 
  summarise(events =  sum(Retirement=="Yes"),time = sum(sum_score)) %>% 
  dplyr::select(-c(1)) %>% 
  as.matrix
retirement_rate <- retirement_rate[c(2,1),]
epi.2by2(retirement_rate, method = "cohort.time")


# Option 3: Manual

# SURFACE (GAMES) --------------------------------------------------------------

# Comparing `surface` vs. `Retirement`
ExpCustomStat(dades,Cvar=c("surface","Retirement"),gpby=TRUE,filt=NULL)
(retSurface<-table(dades$Retirement, dades$surface))

totalgamesSurface<-aggregate(sum_score~surface, data = dades, sum)
totalgamesSurface

# Comparing Clay vs. Carpet
Clay_Carpet<-c(67,67669,10,19685)
epi.2by2(dat = Clay_Carpet, method = "cohort.time", digits = 2,
conf.level = 0.95, units = 1000, interpret = FALSE, outcome = "as.columns")

# Comparing Grass vs. Carpet 
Grass_Carpet<-c(6,5494,10,19685)
epi.2by2(dat = Grass_Carpet, method = "cohort.time", digits = 2,
conf.level = 0.95, units = 1000, interpret = FALSE, outcome = "as.columns")

# Comparing Hard vs. Carpet
Hard_Carpet<-c(108,88566,10,19685)
epi.2by2(dat = Hard_Carpet, method = "cohort.time", digits = 2,
conf.level = 0.95, units = 1000, interpret = FALSE, outcome = "as.columns")

# BEST OF (GAMES) --------------------------------------------------------------

# Comparing `surface` vs. `Retirement`
ExpCustomStat(dades,Cvar=c("best_of","Retirement"),gpby=TRUE,filt=NULL)
(retBestof<-table(dades$Retirement, dades$best_of))

totalgamesBestof<-aggregate(sum_score~best_of, data = dades, sum)
totalgamesBestof

# Comparing 5 sets vs. 3 sets
Fivesets_Threesets<-c(165,140028,26,41386)
epi.2by2(dat = Fivesets_Threesets, method = "cohort.time", digits = 2,
conf.level = 0.95, units = 1000, interpret = FALSE, outcome = "as.columns")

# GROUP (GAMES) ----------------------------------------------------------------

# Comparing `surface` vs. `Retirement`
ExpCustomStat(dades,Cvar=c("group","Retirement"),gpby=TRUE,filt=NULL)
(retgroup<-table(dades$Retirement, dades$group))

totalgamesgroup<-aggregate(sum_score~group, data = dades, sum)
totalgamesgroup

# Comparing No Mundial 1 vs. Mundial1
NoMundial1_Mundial1<-c(143,124970,48,56444)
epi.2by2(dat = NoMundial1_Mundial1, method = "cohort.time", digits = 2,
conf.level = 0.95, units = 1000, interpret = FALSE, outcome = "as.columns")

# ROUND (GAMES) ----------------------------------------------------------------

# Comparing `Round` vs. `Retirement`
ExpCustomStat(dades,Cvar=c("round_cat","Retirement"),gpby=TRUE,filt=NULL)
(retRound<-table(dades$Retirement, dades$round_cat))

totalgamesRound<-aggregate(sum_score~round_cat, data = dades, sum)
totalgamesRound

# Comparing Preliminar vs. Final
Preliminar_Final<-c(160,149780,31,31634)
epi.2by2(dat = Preliminar_Final, method = "cohort.time", digits = 2,
conf.level = 0.95, units = 1000, interpret = FALSE, outcome = "as.columns")


# MATCH NUM (GAMES) ------------------------------------------------------------

# Comparing `Match Num` vs. `Retirement`
ExpCustomStat(dades,Cvar=c("match_num","Retirement"),gpby=TRUE,filt=NULL)
(retMatx<-table(dades$Retirement, dades$match_num))

totalgamesMatx<-aggregate(sum_score~match_num, data = dades, sum)
totalgamesMatx

# Comparing 2nd vs. 1st
Segon_Primer<-c(45,52472,38,51755)
epi.2by2(dat = Segon_Primer, method = "cohort.time", digits = 2,
conf.level = 0.95, units = 1000, interpret = FALSE, outcome = "as.columns")

# Comparing 3rd vs. 1st
Tercer_Primer<-c(57,44041,38,51755)
epi.2by2(dat = Tercer_Primer, method = "cohort.time", digits = 2,
conf.level = 0.95, units = 1000, interpret = FALSE, outcome = "as.columns")

# Comparing 4th vs. 1st
Quart_Primer<-c(51,33146,38,51755)
epi.2by2(dat = Quart_Primer, method = "cohort.time", digits = 2,
conf.level = 0.95, units = 1000, interpret = FALSE, outcome = "as.columns")
```

## Interactions

We study interactions by a decision tree.

```{r fig.width=11}
d <- dades
ct <- party::ctree(Retirement~surface+best_of+year+dif_age+match_num+round,
            data = d, controls=party::ctree_control(maxdepth=3))
plot(ct)
```

## Lineality

### Age

Age is not linear as can be seen in the following figures.

```{r}
#-- Count response
f_rate          <- function(x,lab='Yes') sum(x==lab)/sum(!is.na(x))
t_log_age_rate  <- log(tapply(d$Retirement,d$mean_age_cat,f_rate)+.1)
rate            <- as.numeric(t_log_age_rate)
age             <- as.numeric(names(t_log_age_rate))

plot(age,rate,xlab="mean age (years)",ylab="rate",las=1,main="Count data")
lines(lowess(rate~age), col=4,lwd=2)
```

```{r}
##-- Binary response
t_age       <- table(d$mean_age_cat,d$Retirement)
log_odd_age <- log((t_age[,2]+.5) / (t_age[,1]+.5))
age         <- as.numeric(rownames(t_age))

plot(age,log_odd_age,xlab="mean age (years)",ylab="logodd",las=1, main="Binay data")
lines(lowess(log_odd_age~age), col=4,lwd=2)
```

### Year

```{r}
##-- Count response
t_log_year_rate  <- log(tapply(d$Retirement,d$year,f_rate)+.1)
rate             <- as.numeric(t_log_year_rate)
year             <- as.numeric(names(t_log_year_rate))

plot(year,rate,xlab="calendar year",ylab="rate",las=1,main="Count data")
lines(lowess(rate~year), col=4,lwd=2)
```

```{r}
##-- Binary response
t_year       <- table(d$year,d$Retirement)
log_odd_year <- log((t_year[,2]+.5) / (t_year[,1]+.5))
year         <- as.numeric(rownames(t_year))

plot(year,log_odd_year,xlab="mean age (years)",ylab="logodd",las=1, main="Binay data")
lines(lowess(log_odd_year~year), col=4, lwd=2)
```

# Models

## Count data model 

- The logarithm of the number of games is included as offset
- The year is included centered and as a quadratic polynomial
- Two slopes are included for middle age (before and after 27 years)

ANOVA (Type II) of the first additive model:

```{r}
dd <- dades

# Model 0: additive model
model0 <- glm(Retirement_num ~ surface + best_of + year_center + I(year_center^2) +
                               mean_age*mean_age_ind + match_num + round_cat + group, 
              offset=log(sum_score), family = "poisson", data = dd)
Anova(model0)
```

Overdispersion? No

```{r}
dispersiontest(model0,trafo=1)
dispersiontest(model0,trafo=2)
```
Interactions are proposed and a stepwise is made to find a better model according 
to the BIC criterion and then surface and the interaction with age are added.

```{r}
# BIC + surface and mean_age
model4 <- glm(Retirement_num ~ surface + 
                mean_age*mean_age_ind + 
                best_of*match_num,
              offset=log(sum_score),family = "poisson",data = dd)
summary(model4) 

model_count <- model4
```

### Best model according BIC adding surface and interaction with age

Final model:

```{r}
summary(model_count)
```

```{r}
plot(allEffects(model_count))
```

### Forest plot

Next figure shows the incidence Rate Ratio for the different factors in study. 

Regarding the **surfaces**, the rate of retirements in Clay, Hard or Grass 
is more than double than in the carpet surface, being the two formers 
statistically significant.

The behavior of retirements is different depending on the **age** of participants. 
If they are younger (mean age below 27), then the incidence rate (IR) of retirements 
is the half for a 5 years increase. Meanwhile, if the mean age of the players is
at least 27 years, then the IR doubles for 5 years increase, but not statistical
significantly.

Finally, the **number of individual match** also plays a role in the retirements but
it depends on the number of sets for the eliminatory:
- If the matches are at best of 3 sets, retirements are less usual in 3rd and 4th matches.
- Oppositely, if the matches are at best of 5 sets, retirements are more prone to be produced in 3rd and 4th matches.


```{r}
LH <- list(as.numeric(confint(glht(model_count, "surfaceClay  = 0"))$confint),
           as.numeric(confint(glht(model_count, "surfaceGrass = 0"))$confint),                
           as.numeric(confint(glht(model_count, "surfaceHard  = 0"))$confint),
           as.numeric(confint(glht(model_count, "5*mean_age   = 0"))$confint),
           as.numeric(confint(glht(model_count, t(c(rep(0,4),5,rep(0,5),5,rep(0,3)))))$confint), 
           as.numeric(confint(glht(model_count, "match_num2 = 0"))$confint),                  
           as.numeric(confint(glht(model_count, "match_num3 = 0"))$confint), 
           as.numeric(confint(glht(model_count, "match_num4 = 0"))$confint),
           as.numeric(confint(glht(model_count, "match_num2 + best_of5sets:match_num2 = 0"))$confint),    
           as.numeric(confint(glht(model_count, "match_num3 + best_of5sets:match_num3 = 0"))$confint),
           as.numeric(confint(glht(model_count, "match_num4 + best_of5sets:match_num4 = 0"))$confint))  
IRR <- as.data.frame(matrix(exp(unlist(LH)),ncol=3,byrow = TRUE))
names(IRR) <- c('IRR','LL_CI','UL_CI')
IRR$model<- factor(1:nrow(IRR),levels=1:nrow(IRR))
LABS <- data.frame(model= factor(1:(nrow(IRR)+1),levels=(1:(nrow(IRR)+1))),
                   lab1 = c('Subgroup',
                            NA,'All',NA,'Mean age < 27','Mean age >= 27',
                            NA,'Best of 3',NA,NA,'Best of 5',NA),
                   lab2 = c('Comparison',
                            'surface: Clay vs. Carpet','surface: Grass vs. Carpet','surface: Hard vs. Carpet',
                            'age: +5 years','age: +5 years',
                            'match no.: 2 vs. 1','match no.: 3 vs. 1','match no.: 4 vs. 1',
                            'match no.: 2 vs. 1','match no.: 3 vs. 1','match no.: 4 vs. 1'),
                   lab3 = c('IRR',
                            paste0(formatC(IRR$IRR,   format = 'f', digits=2),' (',
                                   formatC(IRR$LL_CI, format = 'f', digits=2),' , ',
                                   formatC(IRR$UL_CI, format = 'f', digits=2),')')))

##-- Middel plot ---------------------------------------------------------------
gg_mid <- ggplot(IRR,aes(y=fct_rev(model))) +
  geom_point(aes(x=IRR), shape=15, size=4, color='darkblue') +
  geom_linerange(aes(xmin=LL_CI, xmax=UL_CI), color='darkblue', linewidth = 1.2) +
  geom_vline(xintercept = 1) +
  geom_vline(xintercept = c(0.1,0.2,0.5,2,5,10), linetype="dashed") +
  geom_hline(yintercept = c(3.5,6.5,8.5), linetype=3) +
  labs(x="Incidence Rate Ratio (IRR)", y="") +
  scale_x_log10(limits=c(0.025,40), breaks= c(0.1,0.2,0.5,2,5,10)) +
  coord_cartesian(ylim=c(1,12), xlim=c(0.1,10)) +
  theme_classic() +
  theme(axis.line.y  = element_blank(),
        axis.ticks.y = element_blank(),
        axis.text.y  = element_blank(),
        axis.title.y = element_blank())
##-- Left plot -----------------------------------------------------------------
gg_left <- ggplot(LABS,aes(y=fct_rev(model))) +
  geom_text(aes(x = 0, label = lab1),
            hjust = 0, size=3.3,
            fontface = if_else(LABS$lab1==LABS$lab1[1],"bold","plain")) +
  geom_text(aes(x = 1.3, label = lab2),
            hjust = 0, size=if_else(LABS$lab2==LABS$lab2[1],3.5,3),
            fontface = if_else(LABS$lab2==LABS$lab2[1],"bold","plain")) +
  coord_cartesian(xlim = c(0, 2.5), ylim = c(1,12)) + 
  theme_void()
##-- Right plot ----------------------------------------------------------------
gg_right <- ggplot(LABS,aes(y=fct_rev(model))) +
  geom_text(aes(x = 1, label = lab3),
            hjust = 0.5, size=if_else(LABS$lab3==LABS$lab3[1],3.5,3),
            fontface = if_else(LABS$lab3==LABS$lab3[1],"bold","plain")) +
  coord_cartesian(xlim = c(0, 2), ylim = c(1,12)) + 
  theme_void()

layout <- c(
  area(t = 0, l = 0, b = 30, r = 5), 
  area(t = 1, l = 6, b = 30, r = 10),
  area(t = 0, l = 11, b = 30, r = 12))

gg_left + gg_mid + gg_right + plot_layout(design = layout)
```


## Logistic Models

### Fitting

The same variables are put in as in counts and the following model is obtained that will be used to calculate the probabilities of withdrawals.

```{r}
# Model 0
form <- model_count$formula
model0 <- glm(form,family = "binomial",data = dd)
model_bin <- model0
summary(model_bin)
```

### Validation

Residuals validate the model.

```{r}
residualPlot(model_bin)
```

### Interpretation

The following plot shows all the effects:

```{r}
plot(allEffects(model_bin))
```

### Probabilities

We use the emmeans package to understand the effects. 

```{r}
emm_surface <- emmeans(model_bin,~surface,              type='response')
emm_age1    <- emmeans(model_bin,~mean_age|mean_age_ind,
                       at=list(mean_age=seq(18,26,2)),  type='response')
emm_age2    <- emmeans(model_bin,~mean_age|mean_age_ind,
                       at=list(mean_age=seq(28,36,2)),  type='response')
emm_match   <- emmeans(model_bin,~match_num|best_of,    type='response')

s_surface <- summary(emm_surface)
s_age1    <- summary(emm_age1)
s_age2    <- summary(emm_age2)
s_match   <- summary(emm_match)

d_emm <- data.frame(Variable = c(rep('Surface',4),
                                 rep('Age',10),
                                 rep('Match no.',8)),
                    Subgroup = c(rep('All',4),
                                 rep('Age <  27',5),
                                 rep('Age >= 27',5),
                                 rep('Best of 3 sets',4),
                                 rep('Best of 5 sets',4)),
                    Category = c(as.character(s_surface$surface),
                                 s_age1$mean_age[1:5],
                                 s_age2$mean_age[1:5],
                                 s_match$match_num),
                    prob     = c(s_surface$prob,
                                 s_age1$prob[1:5],
                                 s_age2$prob[6:10],
                                 s_match$prob[1:4],
                                 s_match$prob[5:8]),
                    lower    = c(s_surface$asymp.LCL,
                                 s_age1$asymp.LCL[1:5],
                                 s_age2$asymp.LCL[6:10],
                                 s_match$asymp.LCL[1:4],
                                 s_match$asymp.LCL[5:8]),
                    upper    = c(s_surface$asymp.UCL,
                                 s_age1$asymp.UCL[1:5],
                                 s_age2$asymp.UCL[6:10],
                                 s_match$asymp.UCL[1:4],
                                 s_match$asymp.UCL[5:8]))

kable(d_emm, digits=3)
```


